Competing Constructions for German Measure Noun Phrases: from Usage Data to Experiment

### 2017-12-19 Latex/article.tex
More comments from reviewers implemented.

The present study is conducted within the general paradigm of cognitive corpus linguistics and includes a comparison of the corpus findings with results from two experiments.
I assume a model of similarity-based classification in the form of Prototype Theory for modeling alternation.
Protoype Theory has been used in alternation research in cognitively oriented corpus linguistics (see \citealp{DivjakArppe2013}, \citealp{Gries2003}).
It is assumed that the variables annotated in a corpus study on an alternation phenomenon define prototypes for the alternants.
The similarity of a given feature vector (in a concrete sentence where one of the alternants is used) influences to a large extent which alternant is chosen by speakers.
This entails that a variant of Prototype Theory with features is assumed \citep{Rosch1978}, instead of the monolithic prototypes of earlier versions.
Prototype Theory is well suited for modeling constructional choices but it is just one of several similarity-based theories of classification, the most prominent other framework being Exemplar Theory (\citealp{MedinSchaffer1978,Hintzman1986}; see \citealp{StormsEa2000} for a comparison of the theories in experimental settings).
Prototype Theory and Exemplar Theory model essentially the same types of effects and differ mainly in whether they assume higher-level abstractions as part of the cognitive representation (Prototype Theory) or try to do without them (Exemplar Theory).
I am sceptical that corpus analysis alone could ever decide which theory is more suitable, given that substantial doubt has been voiced whether even experimental methods are ultimately able to do so \citep{Barsalou1990}.%
\footnote{In \cite{DivjakArppe2013}, it was shown using corpus data how both models form a converging picture.}
Prototype Theory (with feature abstractions) is preferred here simply because it fits the established alternation modeling paradigm, which relies on features being weighted in statistical models.
In Section~\ref{sec:analyses}, a prototype-theoretical parlance is therefore adopted.

@@

Second, therefore, I focus on grammaticalisation.
It is often assumed that pseudo-partitives arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages in general).
%
%\footnote{Notice that the description of German in \citet[547--549]{Koptjevskaja2001} is vague and partly incorrect, maybe because her main source \cite{Eschenbach1993} is not as clear as it could be (see also Footnote \ref{fn:eschenbash} above).
%Her example (29b) from \citet[71]{Eschenbach1993}, repeated here verbatim as (\ref{ex:koptjewskaja}) is misclassified as a clear case of a dative on the measure noun and a genitive on the kind noun (hence, PGC).
%
%\begin{exe}
%  \ex{\label{ex:koptjewskaja}\gll trotz einer Flasche gut-en Wein-s\\
%  in.spite.of one-\textsc{fem.gen} bottle good-\textsc{masc.sg.gen} wine-\textsc{gen}\\
%  \glt in spite of one bottle of good wine}
%\end{exe}
%
%It could, however, equally well be a genitive on the measure noun because of the syncretisms discussed here on p.\ \pageref{page:femininesyncretism} and an ambiguous case requirement of the preposition \textit{troz}, hence an instance of NAC.
%Then, she does not clearly distinguish acceptable from non-acceptable constructions (cf.\ Section \ref{sec:descriptive}) on her pp.\ 548 and 549, including a rather confusing \textit{???} rating of her (31b), which is clearly a non-acceptable construction of the type illustrated in (\ref{ex:analyticalpartitive:b}) above.
%}
%
The grammaticalisation paths uncovered by \citet[esp.\ 526--530]{Koptjevskaja2001} are relevant for the case at hand.
The grammaticalisation path can start out (in some languages) with constructions involving two referential nouns (not necessarily forming a single and contiguous NP) and a \textit{separative} meaning as in \textit{(cut) two slices from the cake} \citep[535]{Koptjevskaja2001}.
The \textit{part-of} meaning of true partitives as in \textit{a slice of the cake} represents the first stage of a development wherein the measure noun can already lose some semantic content, when for example words like \textit{bite} are no longer necessarily interpreted as a piece literally bitten out of something.
The pseudo-partitive stage finally instantiates a \textit{quantity-of} relation, potentially even leading to fully grammaticalised quantifiers such as \textit{a lot}.
In German, the PGC is clearly the older construction \citep{Zimmer2015}.
It still has the potential to form a true partitive (if the kind noun is definite).
Conversely, the NAC lacks this ability to form true partitives, and it thus forms the more prototypical environment for expressing measurement (as opposed to partitivity).
Hence, we can expect the \NACa\ to be a prototypical hosting construction for more strongly grammaticalised measure nouns.
For example, highly grammaticalised non-referential nouns like \textit{Gramm} `gram' and \textit{Meter} `metre' should occur proportionally more often in the \NACa\ than in the \PGCa.
% \footnote{In case studies like \cite{Brems2003,DeclerckBrems2016} syntactic evidence for grammaticalisation plays a prominent role.
% In English -- \cite{Brems2003} argues based on a view from \cite{Langacker1991} -- strongly lexicalised and thus semantically bleached measure nouns turn from nouns into quantifiers, leading to a syntactic reanalysis of the NP, wherein the kind noun becomes the head.
% The argument is supported by convincing syntactic evidence, for example subject-verb agreement.
% As argued above, in German MNPs the measure noun always remains the syntactic head, and there is no similarly handy criterion to diagnose a high degree of grammaticalisation.}

@@

Third, as described above, the grammaticalisation path leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
Both types of relations can be numerically quantified -- inasmuch as a precise number of \textit{parts} or a numerically exact \textit{quantity} can be specified.
However, it is much more prototypical of quantities to be specified with numerical precision.
This is most obviously so with the highly grammaticalised physical measure nouns like \textit{centilitre} which are very prototypically used with exact numeral instead of unspecific quantifiers, although both options are available (\textit{three centilitres} vs.\ \textit{several centilitres}).
Since the \NACa\ is more closely associated with the \textit{quantity-of} relation, cardinals as attributes of the measure noun are expected to have a higher proportional frequency in the \NACa.
For illustration, (\ref{ex:determiners}) shows the expected alternants under this hypothesis.

@@

Fourth and finally, it was found that the \PGCa\ is more typical of higher stylistic levels (distinctly edited, closer to the non-regional standard, more formal) and\slash or even exclusive to written language (see \citealp[320--323]{Hentschel1993}).
This is not surprising, given that the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial vernacular variants of German as a result of a diachronic process wherein many (but by no means all) uses of the genitive are replaced by other cases or periphrastic constructions \citep{FleischerSchallert2011}.
Under an integral view of prototypes, which incorporate effects related to larger contexts, registers, and styles, such preferences can be part of what defines the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, stylistic effects will therefore be modelled -- even if only using two very simple proxy variables.

@@

For the present study, I used the German \textit{Corpus from the Web} (COW) in its 2014 version DECOW14A (\citealp{SchaeferBildhauer2012full}, and \citealp{Schaefer2015b}, as well as \citealp{BiemannEa2013}, and \citealp{SchaeferBildhauer2013}, for overviews of web corpora in general and the methodology of their construction), which contains almost 21 billion tokens.%
\footnote{The COW corpora (Dutch, English, French, German, Spanish, Swedish) are made available for free at \url{https://www.webcorpora.org}.
At the time of this writing, a newer 2016 version DECOW16 has already been released.}
I chose this corpus for two main reasons.%
\footnote{The use of web data for linguistic research does require explicit and careful justification.
Due to the noisy nature and unknown composition of the web, only carefully designed and established web corpora like the COW corpora or the SketchEngine corpora \citep{KilgarriffEa2014} should be used.
Clearly, using search engine results is ``bad science'' for many reasons, most prominently total irreproducibility of results, as \cite{Kilgarriff2006} pointed out more than ten years ago.
Careless use of search engine results is still found, however, see for example \citet[171--175]{DeclerckBrems2016}.}
First, the external validity of any study is increased through a higher heterogeneity of the sample \citep[30]{MaxwellDelaney2004}, and the DECOW corpus has clearly a much more heterogeneous composition compared to the only other very large corpus of German, the DeReKo \citep{KupietzEa2010} of the Institute for the German Language (IDS), which contains almost exclusively newspaper texts.%
\footnote{It was shown in \cite{W16-2601} that, for example, the spread of topics is much smaller in DeReKo compared to DECOW.}
Second, it was already mentioned that normative grammars often adopt clear positions regarding the grammaticality of either the \NACa\ or the \PGCa.
Thus, newspaper text or any other text that conforms strongly to normative grammars might not represent the alternation phenomenon fully (and without bias) because authors and proofreaders who must adhere to normative guidelines might favour one alternative or the other explicitly.
Web corpora, on the other hand, contain at least some amount of non-standard language from forums and similar sources.
For these or similar reasons, COW corpora have been used in a number of peer-reviewed publications, for example \cite{VanGoethemHiligsmann2014}, \cite{VanGoethemHuening2015}, \cite{MuellerS2014}, \cite{Schaefer2016c}, \cite{SchaeferSayatz2014}, \cite{SchaeferSayatz2016}, and \cite{Zimmer2015}. 
Therefore, DECOW is the obvious choice for this study.

@@

This resulting list of mass nouns was used in step (\ref{enum:data:step2}) to derive a list of measure nouns co-occurring with the mass nouns. 
In order to generate this list, I utilised the fact that a direct sequence of two nouns almost always instantiates the bare-noun NAC if the second noun is a mass noun.
Hence, I searched for all sequences N\Sub{1}N\Sub{2} where N\Sub{2} was one of the mass noun lemmas extracted in step (\ref{enum:data:step1}).
Then, the resulting 100 lists of noun-noun combinations were each sorted by frequency in descending order and sieved manually to remove erroneous hits.
From each of the 100 lists, I also removed noun-noun combinations that had a frequency below 2, except if the individual list would have otherwise been shorter than 20 noun-noun combinations.
The result was a list of the most frequent 2,365 individual combinations of a measure noun and a mass noun.

In step (\ref{enum:data:step3}), each of these 2,365 noun–noun combinations was queried in the target constructions (\PGCa\ and \NACa) individually in each of the first ten slices of DECOW (roughly 10 billion tokens).
In order to reduce the sample size for the manual annotation process, the final concordance was sampled from the results of these 2,365 queries.
Since the mass nouns in the sample were distributed according to the usual power law (often referred to as a \textit{Zipfian} distribution), I used all hits for nouns with a frequency up to 100 and a sample of 100 of all those with higher frequency.
The final sample contained 6,843 sentences, which was reduced to 5,063 in the manual annotation process due to removal of noisy material, erroneous hits and uninformative cases where the measure noun was in the genitive, in which case the \NACa\ cannot be distinguished from the \PGCa.
Given the careful sampling procedure described in this section, we can be highly sure that it contains all relevant and reasonably frequent noun–noun combinations in the target constructions.%
\footnote{In a similar fashion, the 100 most frequent measure nouns occurring with plural kind nouns were listed and queried, resulting in a sample of 871 sentences.
As stated in Section~\ref{sec:germanmeasurenps}, the \NACa\ is virtually never used with plural kind nouns, and this sample was not used except for quantifying the frequency of occurrence of the constructions (67 times \NACa\ and 794 times \PGCa).
However, the sample is distributed in the data package accompanying this paper.
}

@@

In this section, I justify the choice of statistical models which I use in Section~\ref{sec:corpushierarchicalmodel}.
Readers might think that the method used here for modeling grammatical alternations -- namely (Hierarchical) Logistic Regression\slash Generalised Linear (Mixed) Models or GL(M)Ms -- does not require much justification.
After all, GLMMs have been established as the major tool in the analysis of alternation phenomena.
All studies mentioned at the outset of Section~\ref{sec:cogocl} use some form of regression\slash GLMM.
Over the past few years, however, modified or alternative methods have been proposed.
From among these methods, I just make a few remarks on Bayesian estimation (see \citealp{GelmanEa2014}) as it was proposed in \cite{Levshina2016} and \cite{Divjak2016a}, for example.
Conceptually, I see three points of discussion that should be kept apart.
First, Bayesian methods are sometimes touted as superior tools for scientific inference compared to frequentist methods.
Second, it has been proposed that the Bayesian interpretation of probability is more cognitively adequate for the modeling of linguistic data \cite[301--302]{Divjak2016a}.
Third and very specific to this paper, given established methods in the modeling of alternation and variation, it has to be decided whether so-called Bayesian methods lead to substantially different results.

@@

\begin{figure}[hb!]
  \centering
  \includegraphics[width=0.85\textwidth]{../R/output/corpus_fixeffs_mle+mcmc}
  \caption{Coefficients (MLE and MCMC) with 95\% confidence intervals (for details see text); the intercept (\textit{Cardinal=Yes}, \textit{Measurecase=Nom}, \textit{Kindgender=Masc}, \textit{Measureclass=Physical}; 0 for all numeric z-transformed regressors) is -4.328 (MLE) and -4.441 (MCMC)}
  \label{fig:fixeffs}
\end{figure}

In this section, I report the results of fitting a multilevel model to the data using R \citep{R}, \textit{lme4} \citep{lme4} for Maximum Likelihood Estimation (MLE) and \textit{rstanarm} \citep{rstanarm} for Bayesian Markov-Chain Monte Carlo (MCMC) estimation (see Section~\ref{sec:rightstatistics}).
The purpose is to model the influence of the regressors specified in Table~\ref{tab:variables} on the probability that the \PGCa\ is chosen over the \NACa.
All regressors from Table~\ref{tab:variables} were included, and the measure lemma and the kind noun lemma were specified as varying-intercept random effects.
The sample size was \textit{n=}5,063 with 1,134 cases of \PGCa\ and 3,929 cases of \NACa.
The results of the estimation are shown in Table~\ref{tab:bigtable} and in Figure~\ref{fig:fixeffs}.
The regressors with the measure lemma as their unit of reference have no within-measure lemma variance, and the \textit{glmer} function automatically estimates them as \textit{group level predictors} (or \textit{second-level effects}), cf.\ \citet[265--269,302--304]{GelmanHill2006}.
The same goes for those listed with the kind lemma as their unit of reference.
Given the coding of the response variable, coefficients leaning to the positive side can be interpreted as favouring the \PGCa.

\begin{sidewaystable}
  \centering
  \resizebox{\textheight}{!}{
  \begin{tabular}{llrlp{0.5em}rrp{0.5em}rrp{0.5em}rrp{0.5em}cc}
      Level & Regressor & \multicolumn{1}{l}{\pPB} & Level  && \multicolumn{2}{l}{Coefficient} && \multicolumn{2}{l}{CI low} && \multicolumn{2}{l}{CI high} && \multicolumn{2}{l}{CI excludes 0} \\
              &                   &       &           && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} \\\midrule
       First     & Badness           &  0.002 &           && -0.152 & -0.155 && -0.247 & -0.247 && -0.061 & -0.065 && * & * \\
                 & Cardinal          &  0.001 & No        &&  1.189 &  1.222 &&  0.862 &  0.927 &&  1.466 &  1.496 && * & * \\
                 & Genitives         &  0.001 &           && -0.693 & -0.711 && -0.768 & -0.801 && -0.592 & -0.616 && * & * \\
                 & Measurecase       &  0.001 & Acc       &&  0.030 &  0.031 && -0.150 & -0.159 &&  0.212 &  0.222 &&   &   \\
                 &                   &        & Dat       &&  0.705 &  0.729 &&  0.455 &  0.465 &&  0.944 &  0.995 && * & * \\[0.5\baselineskip]

@@

  \end{tabular}
  }
  \caption{Coefficient table comparing Maximum Likelihood Estimation (MLE, with 95\% bootstrap confidence interval) and `Bayesian' Markov-Chain Monte Carlo estimation (MCMC); the intercept (\textit{Cardinal=Yes}, \textit{Measurecase=Nom}, \textit{Kindgender=Masc}, \textit{Measureclass=Physical}; 0 for all numeric z-transformed regressors) is -3.548 (MLE) and -3.700 (MCMC)}
  \label{tab:bigtable}
\end{sidewaystable}

@@

The coefficient estimates are specified in Table~\ref{tab:bigtable} for each regressor (or regressor level) in the columns labelled \textit{Coefficient}.
For a robust quantification of the precision of the estimation, I ran a parametric bootstrap (using the \mbox{\textit{confint.merMod}} function from \textit{lme4}) with 1,000 replications and using the percentile method for the calculation of the intervals.
The resulting 95\% bootstrap confidence intervals are reported in Table~\ref{tab:bigtable} in the columns labelled \textit{CI low} and \textit{CI high} (= upper and lower 2.5th percentiles).
The column \textit{CI excludes 0} shows an asterisk for those intervals that do not include 0.
Furthermore, for each regressor, a p-value was obtained by dropping the regressor from the full model, re-estimating the nested model, and comparing it to the full model.
Instead of inexact Wald approximations and Likelihood Ratio Tests, I used a drop-in bootstrap replacement for the Likelihood Ratio Test from the function \textit{PBmodcomp} from the \textit{pbkrtest} package \citep{HalekohHojsgaard2014}.
I call the corresponding value $p_{\text{PB}}$\ and it is given in the appropriately labelled columns in Table~\ref{tab:bigtable}.
Only \textit{Kindfreq} ($\mpPB=0.095$) can be seen as slightly too high to be convincing (non-significant).

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.75\textwidth]{../R/output/corpus_Measureclass}
  \caption{Effect plot for the regressor \textit{Measureclass}}
  \label{fig:eff:measureattraction}
\end{figure}

>>>>

The present study is conducted within the general paradigm of cognitive corpus linguistics and includes a comparison of the corpus findings with results from two experiments.
I assume a model of similarity-based classification in the form of Prototype Theory for modeling alternation.
Protoype Theory has been used in alternation research in cognitively oriented corpus linguistics (see \citealp{DivjakArppe2013}; \citealp{Gries2003}).
It is assumed that the variables annotated in a corpus study on an alternation phenomenon define prototypes for the alternants.
The similarity of a given feature vector (in a concrete sentence where one of the alternants is used) influences to a large extent which alternant is chosen by speakers.
This entails that a variant of Prototype Theory with features is assumed \citep{Rosch1978}, instead of the monolithic prototypes of earlier versions.
Prototype Theory is well suited for modeling constructional choices but it is just one of several similarity-based theories of classification, the most prominent other framework being Exemplar Theory (\citealp{MedinSchaffer1978,Hintzman1986}; see \citealp{StormsEa2000} for a comparison of the theories in experimental settings).
Prototype Theory and Exemplar Theory model essentially the same types of effects and differ mainly in whether they assume higher-level abstractions as part of the cognitive representation (Prototype Theory) or try to do without them (Exemplar Theory).
I am sceptical that corpus analysis alone could ever decide which theory is more suitable, given that substantial doubt has been voiced whether even experimental methods are ultimately able to do so \citep{Barsalou1990}.%
\footnote{In \cite{DivjakArppe2013}, it was shown using corpus data how both models form a converging picture.}
Prototype Theory (with feature abstractions) is preferred here simply because it fits the established alternation modeling paradigm, which relies on features being weighted in statistical models.
In Section~\ref{sec:analyses}, a prototype-theoretical parlance is therefore adopted.

@@

Second, therefore, I focus on grammaticalisation.
It is often assumed that pseudo-partitives arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages in general).
%
%\footnote{Notice that the description of German in \citet[547--549]{Koptjevskaja2001} is vague and partly incorrect, maybe because her main source \cite{Eschenbach1993} is not as clear as it could be (see also Footnote \ref{fn:eschenbash} above).
%Her example (29b) from \citet[71]{Eschenbach1993}, repeated here verbatim as (\ref{ex:koptjewskaja}) is misclassified as a clear case of a dative on the measure noun and a genitive on the kind noun (hence, PGC).
%
%\begin{exe}
%  \ex{\label{ex:koptjewskaja}\gll trotz einer Flasche gut-en Wein-s\\
%  in.spite.of one-\textsc{fem.gen} bottle good-\textsc{masc.sg.gen} wine-\textsc{gen}\\
%  \glt in spite of one bottle of good wine}
%\end{exe}
%
%It could, however, equally well be a genitive on the measure noun because of the syncretisms discussed here on p.\ \pageref{page:femininesyncretism} and an ambiguous case requirement of the preposition \textit{troz}, hence an instance of NAC.
%Then, she does not clearly distinguish acceptable from non-acceptable constructions (cf.\ Section \ref{sec:descriptive}) on her pp.\ 548 and 549, including a rather confusing \textit{???} rating of her (31b), which is clearly a non-acceptable construction of the type illustrated in (\ref{ex:analyticalpartitive:b}) above.
%}
%
The grammaticalisation paths uncovered by \citet[esp.\ 526--530]{Koptjevskaja2001} are relevant for the case at hand.
The grammaticalisation path can start out (in some languages) with constructions involving two referential nouns (not necessarily forming a single and contiguous NP) and a \textit{separative} meaning as in \textit{(cut) two slices from the cake} \citep[535]{Koptjevskaja2001}.
The \textit{part-of} meaning of true partitives as in \textit{a slice of the cake} represents the first stage of a development wherein the measure noun can already lose some semantic content, when for example words like \textit{bite} are no longer necessarily interpreted as a piece literally bitten out of something.
The pseudo-partitive stage finally instantiates a \textit{quantity-of} relation, potentially even leading to fully grammaticalised quantifiers such as \textit{a lot}.
In German, the PGC is clearly the older construction \citep{Zimmer2015}.
It still has the potential to form a true partitive (if the kind noun is definite).
Conversely, the NAC lacks this ability to form true partitives, and it thus forms the more typical environment for expressing measurement (as opposed to partitivity).
Hence, we can expect the \NACa\ to be a typical hosting construction for more strongly grammaticalised measure nouns.
For example, highly grammaticalised non-referential nouns like \textit{Gramm} `gram' and \textit{Meter} `metre' should occur proportionally more often in the \NACa\ than in the \PGCa.
% \footnote{In case studies like \cite{Brems2003,DeclerckBrems2016} syntactic evidence for grammaticalisation plays a prominent role.
% In English -- \cite{Brems2003} argues based on a view from \cite{Langacker1991} -- strongly lexicalised and thus semantically bleached measure nouns turn from nouns into quantifiers, leading to a syntactic reanalysis of the NP, wherein the kind noun becomes the head.
% The argument is supported by convincing syntactic evidence, for example subject-verb agreement.
% As argued above, in German MNPs the measure noun always remains the syntactic head, and there is no similarly handy criterion to diagnose a high degree of grammaticalisation.}

@@

Third, as described above, the grammaticalisation path leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
Both types of relations can be numerically quantified -- inasmuch as a precise number of \textit{parts} or a numerically exact \textit{quantity} can be specified.
However, it is much more typical of quantities to be specified with numerical precision.
This is most obviously so with the highly grammaticalised physical measure nouns like \textit{centilitre} which are very typically used with exact numeral instead of unspecific quantifiers, although both options are available (\textit{three centilitres} vs.\ \textit{several centilitres}).
Since the \NACa\ is more closely associated with the \textit{quantity-of} relation, cardinals as attributes of the measure noun are expected to have a higher proportional frequency in the \NACa.
For illustration, (\ref{ex:determiners}) shows the expected alternants under this hypothesis.

@@

Fourth and finally, it was found that the \PGCa\ is more typical of higher stylistic levels (distinctly edited, closer to the non-regional standard, more formal) and\slash or even exclusive to written language (see \citealp[320--323]{Hentschel1993}).
This is not surprising, given that the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial vernacular variants of German as a result of a diachronic process wherein many (but by no means all) uses of the genitive are being replaced by other cases or periphrastic constructions \citep{FleischerSchallert2011}.
Under an integral view of prototypes, which incorporate effects related to larger contexts, registers, and styles, such preferences can be part of what defines the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, stylistic effects will therefore be modelled -- even if only using two very simple proxy variables.

@@

For the present study, I used the German \textit{Corpus from the Web} (COW) in its 2014 version DECOW14A (\citealp{SchaeferBildhauer2012full}, and \citealp{Schaefer2015b}, as well as \citealp{BiemannEa2013}, and \citealp{SchaeferBildhauer2013}, for overviews of web corpora in general and the methodology of their construction), which contains almost 21 billion tokens.%
\footnote{The COW corpora (Dutch, English, French, German, Spanish, Swedish) are made available for free at \url{https://www.webcorpora.org}.
At the time of this writing, a newer 2016 version DECOW16 has already been released.}
I chose this corpus for two main reasons.%
\footnote{The use of web data for linguistic research does require explicit and careful justification.
Due to the noisy nature and unknown composition of the web, only carefully designed and established web corpora like the COW corpora or the SketchEngine corpora \citep{KilgarriffEa2014} should be used.
Clearly, using search engine results is ``bad science'' for many reasons, most prominently total non-replicability of results, as \cite{Kilgarriff2006} pointed out more than ten years ago.
Careless use of search engine results is still found, however, see for example \citet[171--175]{DeclerckBrems2016}.}
First, the external validity of any study is increased through a higher heterogeneity of the sample \citep[30]{MaxwellDelaney2004}, and the DECOW corpus has clearly a much more heterogeneous composition compared to the only other very large corpus of German, the DeReKo \citep{KupietzEa2010} of the Institute for the German Language (IDS), which contains almost exclusively newspaper texts.%
\footnote{It was shown in \cite{W16-2601} that, for example, the range of topics covered is much smaller in DeReKo compared to DECOW.}
Second, it was already mentioned that normative grammars often adopt clear positions regarding the grammaticality of either the \NACa\ or the \PGCa.
Thus, newspaper text or any other text that conforms strongly to normative grammars might not represent the alternation phenomenon fully (and without bias) because authors and proofreaders who must adhere to normative guidelines might favour one alternative or the other explicitly.
Web corpora, on the other hand, contain at least some amount of non-standard language from forums and similar sources.
For these or similar reasons, COW corpora have been used in a number of peer-reviewed publications, for example \cite{VanGoethemHiligsmann2014}, \cite{VanGoethemHuening2015}, \cite{MuellerS2014}, \cite{Schaefer2016c}, \cite{SchaeferSayatz2014}, \cite{SchaeferSayatz2016}, and \cite{Zimmer2015}. 
Therefore, DECOW is a valid choice for this study.

@@

This list of mass nouns was used in step (\ref{enum:data:step2}) to derive a list of measure nouns co-occurring with the mass nouns. 
In order to generate this list, I utilised the fact that a direct sequence of two nouns almost always instantiates the bare-noun NAC if the second noun is a mass noun.
Hence, I searched for all sequences N\Sub{1}N\Sub{2} where N\Sub{2} was one of the mass noun lemmas extracted in step (\ref{enum:data:step1}).
Then, the resulting 100 lists of noun-noun combinations were each sorted by frequency in descending order and sieved manually to remove erroneous hits.
From each of the 100 lists, I also removed noun-noun combinations that had a frequency below 2, except if the individual list would have otherwise been shorter than 20 noun-noun combinations.
The result was a list of the most frequent 2,365 individual combinations of a measure noun and a mass noun.

In step (\ref{enum:data:step3}), each of these 2,365 noun–noun combinations was queried in the target constructions (\PGCa\ and \NACa) individually in each of the first ten slices of DECOW (roughly 10 billion tokens).
In order to reduce the sample size for the manual annotation process, the final concordance was sampled from the results of these 2,365 queries.
Since the mass nouns in the sample were distributed according to the usual power law (often referred to as a \textit{Zipfian} distribution), I used all hits for nouns with a frequency up to 100 and a sample of 100 of all those with higher frequency.
The final sample contained 6,843 sentences, which was reduced to 5,063 in the manual annotation process due to removal of noisy material, erroneous hits and uninformative cases where the measure noun was in the genitive, in which case the \NACa\ cannot be distinguished from the \PGCa.
Given the careful sampling procedure described in this section, we can be highly sure that it contains all relevant and reasonably frequent noun–noun combinations in the target constructions.%
\footnote{In a similar fashion, the 100 most frequent measure nouns occurring with plural kind nouns were listed and queried, resulting in a sample of 871 sentences.
As stated in Section~\ref{sec:germanmeasurenps}, the \NACa\ is virtually never used with plural kind nouns, and this sample was not used except for quantifying the frequency of occurrence of the constructions (67 times \NACa\ and 794 times \PGCa).
However, the sample is distributed with the data package accompanying this paper.
}

@@

In this section, I justify my choice of statistical models used in Section~\ref{sec:corpushierarchicalmodel}.
Readers might think that the method used here for modeling grammatical alternations -- namely (Hierarchical) Logistic Regression\slash Generalised Linear (Mixed) Models or GL(M)Ms -- does not require much justification.
After all, GLMMs have been established as the major tool in the analysis of alternation phenomena.
All studies mentioned at the outset of Section~\ref{sec:cogocl} use some form of regression\slash GLMM.
Over the past few years, however, modified or alternative methods have been proposed.
While it is impossible to discuss all of these methods here, I just make a few remarks on Bayesian estimation (see \citealp{GelmanEa2014}) as it was proposed in \cite{Levshina2016} and \cite{Divjak2016a}, for example.
Conceptually, I see three points of discussion that should be kept apart.
First, Bayesian methods are sometimes touted as superior tools for scientific inference compared to frequentist methods.
Second, it has been proposed that the Bayesian interpretation of probability is more cognitively adequate for the modeling of linguistic data \cite[301--302]{Divjak2016a}.
Third and very specific to this paper, given established methods in the modeling of alternation and variation, it has to be decided whether so-called Bayesian methods lead to substantially different results.

@@

\begin{figure}[hb!]
  \centering
  \includegraphics[width=0.85\textwidth]{../R/output/corpus_fixeffs_mle+mcmc}
  \caption{Coefficients (MLE and MCMC) with 95\% confidence intervals (for details see text); the intercept is -4.328 (MLE) and -4.441 (MCMC)}
  \label{fig:fixeffs}
\end{figure}

In this section, I report the results of fitting a multilevel model to the data using R \citep{R}, \textit{lme4} \citep{lme4} for Maximum Likelihood Estimation (MLE) and \textit{rstanarm} \citep{rstanarm} for Bayesian Markov-Chain Monte Carlo (MCMC) estimation (see Section~\ref{sec:rightstatistics}).
The purpose is to model the influence of the regressors specified in Table~\ref{tab:variables} on the probability that the \PGCa\ is chosen over the \NACa.
All regressors from Table~\ref{tab:variables} were included, and the measure lemma and the kind noun lemma were specified as varying-intercept random effects.
The sample size was \textit{n=}5,063 with 1,134 cases of \PGCa\ and 3,929 cases of \NACa.
The results of the estimation are shown in Table~\ref{tab:bigtable} and in Figure~\ref{fig:fixeffs}.
The intercept comprises \textit{Cardinal=Yes}, \textit{Measurecase=Nom}, \textit{Kindgender=Masc}, \textit{Measureclass=Physical}, and 0 for all numeric z-transformed regressors.
It is -4.328 for the ML estimate and -4.441 for the MCMC estimate.

The regressors with the measure lemma as their unit of reference have no within-measure lemma variance, and the \textit{glmer} function automatically estimates them as \textit{group level predictors} (or \textit{second-level effects}), cf.\ \citet[265--269,302--304]{GelmanHill2006}.
The same goes for those listed with the kind lemma as their unit of reference.
Given the coding of the response variable, coefficients leaning to the positive side can be interpreted as favouring the \PGCa.

\begin{sidewaystable}
  \centering
  \resizebox{\textheight}{!}{
  \begin{tabular}{llrlp{0.5em}rrp{0.5em}rrp{0.5em}rrp{0.5em}cc}
      Model & Regressor & \multicolumn{1}{l}{\pPB} & Factor && \multicolumn{2}{l}{Coefficient} && \multicolumn{2}{l}{CI low} && \multicolumn{2}{l}{CI high} && \multicolumn{2}{l}{CI excludes 0} \\
      level &                   &       & level     && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} && \multicolumn{1}{l}{MLE} & \multicolumn{1}{l}{MCMC} \\\midrule
       First     & Badness           &  0.002 &           && -0.152 & -0.155 && -0.247 & -0.247 && -0.061 & -0.065 && * & * \\
                 & Cardinal          &  0.001 & No        &&  1.189 &  1.222 &&  0.862 &  0.927 &&  1.466 &  1.496 && * & * \\
                 & Genitives         &  0.001 &           && -0.693 & -0.711 && -0.768 & -0.801 && -0.592 & -0.616 && * & * \\
                 & Measurecase       &  0.001 & Acc       &&  0.030 &  0.031 && -0.150 & -0.159 &&  0.212 &  0.222 &&   &   \\
                 &                   &        & Dat       &&  0.705 &  0.729 &&  0.455 &  0.465 &&  0.944 &  0.995 && * & * \\[0.5\baselineskip]

@@

  \end{tabular}
  }
  \caption{Coefficient table comparing Maximum Likelihood Estimation (MLE, with 95\% bootstrap confidence interval) and `Bayesian' Markov-Chain Monte Carlo estimation (MCMC); the intercept is -4.328 (MLE) and -4.441 (MCMC)}
  \label{tab:bigtable}
\end{sidewaystable}

@@

The coefficient estimates are specified in Table~\ref{tab:bigtable} for each regressor (or regressor level) in the columns labelled \textit{Coefficient}.
For a robust quantification of the precision of the estimation, I ran a parametric bootstrap (using the \mbox{\textit{confint.merMod}} function from \textit{lme4}) with 1,000 replications and using the percentile method for the calculation of the intervals.
The resulting 95\% bootstrap confidence intervals are reported in Table~\ref{tab:bigtable} in the columns labelled \textit{CI low} and \textit{CI high} (= upper and lower 2.5th percentiles).
The column \textit{CI excludes 0} shows an asterisk for those intervals that do not include 0.
Furthermore, for each regressor, a p-value was obtained by dropping the regressor from the full model, re-estimating the nested model, and comparing it to the full model.
Instead of inexact Wald approximations and Likelihood Ratio Tests, I used a drop-in bootstrap replacement for the Likelihood Ratio Test from the function \textit{PBmodcomp} from the \textit{pbkrtest} package \citep{HalekohHojsgaard2014}.
I call the corresponding value $p_{\text{PB}}$\ and it is given in the respective columns in Table~\ref{tab:bigtable}.
Only \textit{Kindfreq} ($\mpPB=0.095$) can be seen as slightly too high to be convincing (non-significant).

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{../R/output/corpus_Measureclass}
  \caption{Effect plot for the regressor \textit{Measureclass}}
  \label{fig:eff:measureattraction}
\end{figure}
